Variance reduction for MCMC methods via martingale representations

### 2020-03-21 aap/Revision/vr-mcmc-revision-new-new.tex
update

Monte Carlo integration typically has an error variance of the form
$\sigma^{2}/n,$ where $n$ is a sample size and \(\sigma^2\) is the variance of  integrand. We can make the variance smaller
by using a larger value of $n$.  Alternatively,
 we can  reduce $\sigma^2$ instead of increasing the sample size \(n\). To this end, one can try to construct
a new Monte Carlo experiment with the same expectation as the original one
but with a lower variance $\sigma^2$. Methods to achieve this are known as variance
reduction techniques. Variance reduction plays an important role in
Monte Carlo and Markov Chain Monte Carlo methods. Introduction to many of the variance reduction techniques can be found in \cite{christian1999monte}, \cite{rubinstein2016simulation}, \cite{GobetBook} and \cite{glasserman2013monte}. Recently one witnessed a revival of interest in efficient variance reduction methods  for MCMC algorithms, see for example \cite{dellaportas2012control}, \cite{mira2013zero}, \cite{brosse2018diffusion} and references therein.
\par
Suppose that we wish to compute $\pi(f):=\mathsf{E}\left[f(X)\right]$, where $X$
is a random vector-valued in $\mathcal{X}\subseteq\mathbb{R}^{d}$ with a density $\pi$ and $f:$
$\mathcal{X}\to\mathbb{R}$ with $f\in L^2(\pi)$.
The idea of the control variates variance reduction method
 is to find a cheaply computable random variable $\zeta$ with $\mathsf{E}[\zeta]=0$ and \(\mathsf{E}[\zeta^2]<\infty,\)
such that the variance of the r.v. $f(X)-\zeta$ is small.  The complexity of the problem of constructing classes $Z$ of control variates \(\zeta\) satisfying   $\mathsf{E}[\zeta]=0$ essentially depends on the degree of our knowledge on \(\pi.\)
For example, if \(\pi\) is analytically known and satisfies some regularity conditions, one can apply the well-known technique of  polynomial interpolation to construct control variates enjoying  some optimality properties, see, for example, Section~3.2 in \cite{dimov2008monte}. Alternatively, if an orthonormal system in \(L^2(\pi)\) is analytically available, one can build control variates \(\zeta\) as a linear combination of the corresponding basis functions, see \cite{GobetCV}. Furthermore, if \(\pi\) is known only up to a normalizing constant (which is often the case in Bayesian statistics), one can apply the recent approach of control variates  depending only on the gradient \(\nabla \log \pi\)  using Schr\(\ddot{\text{o}}\)dinger-type Hamiltonian operator in  \cite{mira2013zero} and Stein operator in \cite{brosse2018diffusion}.
In some situations \(\pi\) is not known analytically, but \(X\) can be represented as a function of  simple random variables with known distribution.
Such  situation arises, for example, in the case of functionals of  discretized diffusion processes. In this case a Wiener chaos-type decomposition can be used to construct control variates with nice theoretical properties, see \cite{belomestny2018stratified}.
Note that in order to compare different  variance reduction approaches, one has to analyze their complexity, that is, the number of numerical operations required to achieve a prescribed magnitude of the resulting variance.

@@

\begin{cor}
Let \((X^x_{p})_{p\geq 0}\) be a homogeneous Markov chain of the form \eqref{eq:ula_new} with $\Phi_l = \Phi,$ $l\geq 1$. Then for all $p \in \nset,$ $q \leq p$, $j < q \leq p$, $f \in \mathcal{F}_{2,\text{hom}}^p$ and $x \in \rset^d$ it holds in \(L^2(\mathsf{P})\)
\begin{eqnarray*}
f(X^x_{q})=\mathsf{E}\left[\left.f(X^x_{q})\right|\mathcal G_{j}\right]+\sum_{k=1}^{\infty}\sum_{l=j+1}^{q}\bar a_{q-l,k}(X^x_{l-1})\phi_k\left(\xi_{l}\right)
\end{eqnarray*}
where for all $y \in \rset^d$
\begin{eqnarray*}
\bar a_{r,k}(y)=\mathsf{E}\left[f(X^y_{r})\phi_k\left(\xi_{1}\right)\right], \quad r, k\in \mathbb{N}.
\end{eqnarray*}
\end{cor}

@@

\section{Estimation of coefficients}

@@

\widetilde{A}_{q,k} \in \argmin_{\psi \in \Psi} \sum_{s = 1}^{T} \left| \sum_{r = 0}^{q}  f(X_{r+L}^{(s)})\phi_k(\xi^{(s)}_{L+1}) - \psi(X^{(s)}_L)\right|^2

@@

for some $L$ with $L+q\leq N+n,$

@@

\PE[A^2_{l,k}(X^x_{N+l-1})]\leq R^2_k,\quad k=1,2,\ldots, 

@@

for a sequence  $(R_k)$ satisfying $\sum_{k=1}^\infty R^2_k\leq C$ with constant $C$  depending neither on $\gamma$ nor on $n$ (see \eqref{eq:Aqk2}). Hence if we take $U_k=R_k$ as truncation parameters in \eqref{eq:trunc}, then $\widehat A_{n-l,k}$ converges to $A_{n-l,k}$ as $T\to \infty.$ Moreover,

@@

\PVar[\widehat \pi_{K,n}^N(f)|\TrainSet]\leq \frac{2C}{n}, 

@@

for a constant $C$ not depending on $n$ and $\gamma.$
If the class of functions $\Psi$ is linear of dimension $\mathrm{D}_\Psi$ As can be easily seen the cost of computing $\widehat \pi_{K,n}^N(f)$ is of order $ K T^2 n$. At the same time the variance of the standard estimate $\widehat \pi_{n}^N(f)$ is of order $1/(n\gamma)$ and this bound can not be improved in general. Thus the ratio of cost-variances is of order

@@

\frac{\mathrm{cost}(\pi_{K,n}^N) \PVar[\widehat \pi_{K,n}^N(f)|\TrainSet]}{\mathrm{cost}(\pi_{n}^N)\PVar[\pi_{n}^N(f)]}\leq C \gamma K T^2 L^2 \mathrm{Dim}(\Psi). 

@@

If $K,$ $L$ and $\mathrm{Dim}(\Psi)$ are fixed, then our variance reduction method is advantageous if $\gamma=O(1/T^2).$ 

@@

By $H_k$, $k\in\mathbb N_0$,
we denote the normalized Hermite polynomial on $\mathbb R$, that is,
$$
H_k(x)=\frac{(-1)^k}{\sqrt{k!}}\rme^{x^2/2}\frac{\partial^k}{\partial x^k}\rme^{-x^2/2},
\quad x\in\mathbb R.
$$
For a multi-index $\mathbf{k}=(k_i)\in\mathbb N_0^d$,
$\mathbf{H}_\mathbf{k}$ denotes the normalized Hermite polynomial on $\mathbb R^d$, that is,
$$
\mathbf{H}_\mathbf{k}(\mathbf{x})=\prod_{i=1}^d H_{k_i}(x_i),\quad \mathbf{x}=(x_i)\in\mathbb R^d.
$$
In what follows, we also use the notation
$|\mathbf{k}|=\sum_{i=1}^d k_i$ for $\mathbf{k}=(k_i)\in\mathbb N_0^d$,
and we set $\mathcal G_p=\sigma(Z_1,\ldots,Z_p)$, $p\in\mathbb N$, and $\mathcal G_0=\mathrm{triv}$.
Given $N$ and $n$ as above, for $K\in\mathbb N$, denote
\begin{align}
M_{K,n}^N(f) & =\sum_{\mathbf k\colon 0<\|\mathbf{k}\|\le K}\frac{1}{n}\sum_{l=1}^{n}A_{n-l,\mathbf{k}}(X_{N+l-1})\mathbf{H}_\mathbf{k}(Z_{N+l})
\label{eq:29032018a5}
\notag
\end{align}
with \(\|\mathbf{k}\|=\max_{i} k_i\) and 
\begin{eqnarray*}
A_{q,\mathbf{k}}(y)=\sum_{r=0}^{ q } \bar{a}_{r,\mathbf{k}}(y).
\end{eqnarray*}
For an estimator $\rho(f)\in\{\pi_n^N(f),\pi_{K,n}^N(f)\}$
of $\pi(f)$ (see \eqref{eq:29032018a2} and \eqref{eq:29032018a3}), we shall be interested in the Mean Squared Error (MSE),
which can be decomposed as the sum of the squared  bias
and the  variance:
\begin{equation}
\label{eq:29032018a4}
\mathrm{MSE}\left[\rho(f)\right] = \PE\left[\left\{\rho(f)-\pi(f)\right\}^2\right]
=\left\{\PE[\rho(f)]-\pi(f)\right\}^2 +\PVar[\rho(f)].
\end{equation}
Our analysis is carried out under the following two assumptions:
\begin{description}
\item[(H1)][\textbf{Lipschitz continuity}] The potential $U$ is differentiable and $\nabla U$ is Lipschitz, that is, there exists $L < \infty$ such that
\begin{eqnarray*}
\| \nabla U(x)-\nabla U(y) \| \leq L \|x-y\|, \quad x,y\in \rset^d.
\end{eqnarray*}
\item[(H2)][\textbf{Convexity outside a ball}] There exist $K>0$ and $m>0$ such that for any $x\not\in B_{K}(0)$  it holds
\begin{equation*}
\ps{D^2U(x)}{x} \geq m\|x\|^2.
\end{equation*}
\end{description}
Let $\pi$ be the probability measure on $\rset^d$ with density $\pi(x)$ of the form \eqref{eq:stationary_distr};
for $\gamma>0$, define the Markov kernel $Q_\gamma$ associated to one step of the ULA algorithm by
\begin{equation}
\label{eq:ula_kernel}
\kerULA_{\gamma}(x,A)=\int_{A} \frac{1}{(2 \pi \gamma)^{d/2}} \text{ exp} \left\{ -\frac{1}{2\gamma} \left\| y - x + \gamma \mu(x)\right\|^2\right\} \rmd y
\end{equation}
for any $A \in \mathcal{B}(\rset^d)$. Note that for a homogeneous Markov chain $\left(X^x_{p}\right)_{p \in \nset_0}$ of the form \eqref{eq:chain} with constant step size $\gamma$, $\mathsf{P}(X_n^x \in A) = \kerULA^n_{\gamma}(x,A)$.

\paragraph{Squared bias:} Due to the martingale transform structure
of $M_{K,n}^N(f)$,
we have
\[
\mathsf E\left[M_{K,n}^N(f)\right]=0,
\]
Hence both estimators
$\pi_n^N(f)$ and $\pi_{K,n}^N(f)$ have the same conditional bias.

\begin{lem} 
\label{lem:bias}
Let $f$ be a bounded Borel function and $(X^x_{p})_{p \in \nset_0}$ be a Markov chain generated by ULA with constant step size $\gamma$. Then the squared bias satisfies
\begin{equation}
\label{eq:sq_bias}
\left( \mathsf{E}[\rho(f)]-\pi(f)\right)^2 \lesssim \|f\|^2_{\infty}\left(\gamma + \frac{\rho^{2N}}{n(1-\rho^2)}\right) \lesssim \|f\|^2_{\infty}\left(\gamma + \frac{e^{-\gamma \kappa N}}{2\gamma n}\right)
\end{equation}
for $V(x) = 1 + \|x\|^2$, $\rho$ defined in Lemma~\ref{lem:v_ergodicity}, $\kappa$ independent of $N,n,\gamma$ and $\lesssim$ means less up to a constant not depending on $N,n$ and $\gamma$.
\end{lem}

\begin{proof} Under assumptions {\bf (H1)} and {\bf (H2)}, the corresponding Markov chain has a unique stationary distribution $\pi_\gamma$, which is different from $\pi$. Thus we represent the squared bias as 
\begin{eqnarray}
\label{eq:bias_1}
\nonumber
\left( \mathsf{E}[\pi_{K,n}^N(f)]-\pi(f)\right)^2
&=&\left(\mathsf{E}[\pi_{n}^N(f)]-\pi(f)\right)^2
=
\frac{1}{n^2}\left[\sum\limits_{k=N+1}^{N+k}\left(f(X_k^x)-\pi(f)\right)\right]^2 \\
&\leq& 
2|\pi_{\gamma}(f) - \pi(f)|^2 + \frac{2}{n}\sum\limits_{k=N+1}^{N+n}\left[\PE f(X_k^x) - \pi_{\gamma}(f)\right]^2
\end{eqnarray}
 From \cite[Theorem~10]{durmus:moulines:2017} and Lemma~\ref{lem:v_ergodicity} it follows that
\begin{equation}
\label{eq:bias_2}
\tvnorm{\pi - \pi_\gamma} < C_1\sqrt{\gamma}, \quad |\PE f(X_k^x) - \pi_{\gamma}(f)| \leq C_2 \rho^n\|f\|_{V}V(x)
\end{equation}
for  $V(x) = 1 + \|x\|^2,$ $\rho\in (0,1)$ specified in Lemma~\ref{lem:v_ergodicity} and constants $C_1,C_2>0$ independent of $\gamma$  and $n$. Since $f$ is bounded, $\|f\|_V \leq \|f\|_{\infty}$. Hence, combining \eqref{eq:bias_1} and \eqref{eq:bias_2}, we obtain \eqref{eq:sq_bias}.
\end{proof}

\paragraph{Variance:} In this section we derive an upper bound for the variance of the classical estimator~\eqref{eq:29032018a2} for ULA-based chain.

\begin{lem} 
\label{lem:variance}
Let $f$ be a bounded Borel function and $(X^x_{p})_{p \in \nset_0}$ be a Markov chain generated by ULA with constant step size $\gamma$, satisfying {\bf (H1)} and {\bf (H2)} and assume also $n\gamma > C_0$ for some fixed $C_0$. Then it holds

\begin{equation}
\label{eq:var-mc}
\PVar[\pi_n^N(f)]\lesssim \frac{V(x)}{n\gamma} \eqsp,
\end{equation}
with $V(x) = 1+ \|x\|^2$ where $\lesssim$ stands for inequality up to a constant depending on constants $L,$ $m,$  $M$ and $\mathrm{osc}(f)$.
\end{lem}
\begin{proof}
Under assumptions {\bf (H1)} and {\bf (H2)} the Markov chain $(X^x_{p})_{p \in \nset_0}$ is $V-$geometrically ergodic with $V(x) = 1+\|x\|^2$ and $\rho = \rme^{-\kappa\gamma}$ with $\kappa$ specified in lemma~\ref{lem:v_ergodicity}. Due to lemma~\ref{lem:covariance}, 
\[
\PVar[\pi_n^N(f)]\lesssim \frac{1}{n(1-\rho^{1/2})}
\]
and the result of lemma follows from Taylor expansion of denominator.
\end{proof}

One of the main results of this paper is the following  upper bound for the variance of variance-reduced estimator $\pi_{K,n}^{N}(f).$

@@

Suppose additionally that a bounded function $f$ and $\mu=\nabla U$ are $K \times d \geq 2$ times continuously differentiable

@@

Suppose that $n\gamma^2\leq \Gamma.$ Set $V(x) = 1 + \|x\|^2,$ then  it holds

@@

\sum_{\mathbf{k}\colon\|\mathbf{k}\|\geq K} A^2_{q,\mathbf{k}}(x)\leq C_{K}\sum_{I\subseteq\{1,\ldots,d\},\, I\neq \emptyset}

@@

for some constant $C_K$ not depending on $q$ and depending on $\gamma$ only through $\Gamma$. 

@@

\PVar\left(\pi_{K,n}^{N}(f)\right)\lesssim n^{-1} \gamma^{K-1} 

@@

where $\lesssim$ stands for inequality up to a constant depending only on $K$ and $\Gamma$.

@@

A_{q,\mathbf{k}}(x)=\gamma^{|\mathbf{k}'|/2}\frac{\sqrt{(\mathbf{k}-\mathbf{k}')!}}{\sqrt{\mathbf{k}!}}\mathsf{E}_x\left[ \partial_{Z_1}^{\mathbf{k}'} F(x,Z_1,\ldots,Z_q) \mathbf{H}_{\mathbf{k}-\mathbf{k}'}(Z_{1})\right],

@@

stands for a weak partial derivative of the functional $F_s$ that also can be viewed as discretised version of Malliavin derivative. From \eqref{eq:Ask-part} we derive

@@

\sum_{\mathbf{k}\colon\|\mathbf{k}\|\geq K} A^2_{q,\mathbf{k}}(x)\leq \sum_{I\subseteq\{1,\ldots,d\},\, I\neq \emptyset}

@@

where the sum runs over all nonempty subsets $I$ of the set $\{1,\ldots,d\}$ and

@@

\right)\leq C_{K}\gamma^{-1}(1+V(x))

@@

for some constant $C_K$ not depending on $q$ and $\gamma$.

@@

We perform numerical experiments based on ULA algorithm \ref{exam:langevin-algorithm} with constant step size $\gamma$. We construct a polynomial approximation  for each \(Q_r(x)\) in \eqref{eq:definition-bar-a} of the form:

@@

\widehat{Q}_{r}(x) = \sum_{\|\mathbf{s}\|\leq m} \widehat\beta_{\mathbf{s}} x^{\mathbf{s}},\quad s=(s_1,\ldots,s_d) \in \nset_0^d

@@

with some coefficients \(\widehat \beta_{\mathbf{s}}\in \mathbb{R}\) obtained using a modified least-squares criteria based on $T$ independent training trajectories \(\left\{\left(X_1^{(s)},\ldots,X_{N+n}^{(s)}\right)\right\}_{s=1}^{T}\) with $N$ being the size of burn-in period:

@@

\widehat{Q}_r = \argmin_{\psi \in \Psi} \sum_{s = 1}^{T} \sum_{l = N + 1}^{N+n-r} \left| f(X_{l+r}^{(s)}) - \psi(X_l^{(s)})\right|^2

@@

for \(1 \leq r \leq n-1,\) $\Psi$ being the class of polynomials $\Psi = \{\psi(x) | \psi(x) = \sum_{\|\mathbf{s}\|\leq m} \alpha_{\mathbf{s}} x^{\mathbf{s}} \}$ and $\widehat{Q}_0(x) = f(x)$ by definition.

@@

\widehat a_{r,\mathbf{k}} (x) &=&\CPE{\mathbf{H}_\mathbf{k}(x) \widehat{Q}_{r}(x - \gamma \mu(x) + \sqrt{\gamma}\xi)}{\TrainSet \vee \mathcal{G}_N}

>>>>

Monte Carlo integration typically has an error variance of the form
$\sigma^{2}/n,$ where $n$ is a sample size and \(\sigma^2\) is the variance of  integrand. We can make the variance smaller
by using a larger value of $n$.  Alternatively,
 we can  reduce $\sigma^2$ instead of increasing the sample size \(n\). To this end, one can try to construct
a new Monte Carlo experiment with the same expectation as the original one
but with a lower variance $\sigma^2$. Methods to achieve this are known as variance
reduction techniques. Variance reduction plays an important role in
Monte Carlo and Markov Chain Monte Carlo methods. Introduction to many of the variance reduction techniques can be found in \cite{rubinstein2016simulation}, \cite{GobetBook} and \cite{glasserman2013monte}. Recently one witnessed a revival of interest in efficient variance reduction methods  for MCMC algorithms, see for example \cite{dellaportas2012control}, \cite{mira2013zero}, \cite{brosse2018diffusion} and references therein.
\par
Suppose that we wish to compute $\pi(f):=\mathsf{E}\left[f(X)\right]$, where $X$
is a random vector-valued in $\mathcal{X}\subseteq\mathbb{R}^{d}$ with a density $\pi$ and $f:$
$\mathcal{X}\to\mathbb{R}$ with $f\in L^2(\pi)$.
The idea of the control variates variance reduction method
 is to find a cheaply computable random variable $\zeta$ with $\mathsf{E}[\zeta]=0$ and \(\mathsf{E}[\zeta^2]<\infty,\)
such that the variance of the r.v. $f(X)-\zeta$ is small.  The complexity of the problem of constructing classes $Z$ of control variates \(\zeta\) satisfying   $\mathsf{E}[\zeta]=0$ essentially depends on the degree of our knowledge on \(\pi.\)
For example, if \(\pi\) is analytically known and satisfies some regularity conditions, one can apply the well-known technique of  polynomial interpolation to construct control variates enjoying  some optimality properties, see, for example, Section~3.2 in \cite{dimov2008monte}. Alternatively, if an orthonormal system in \(L^2(\pi)\) is analytically available, one can build control variates \(\zeta\) as a linear combination of the corresponding basis functions, see \cite{GobetCV}. Furthermore, if \(\pi\) is known only up to a normalizing constant (which is often the case in Bayesian statistics), one can apply the recent approach of control variates  depending only on the gradient \(\nabla \log \pi\)  using Schr\(\ddot{\text{o}}\)dinger-type Hamiltonian operator in  \cite{mira2013zero} and Stein operator in \cite{brosse2018diffusion}.
In some situations \(\pi\) is not known analytically, but \(X\) can be represented as a function of  simple random variables with known distribution.
Such  situation arises, for example, in the case of functionals of  discretized diffusion processes. In this case a Wiener chaos-type decomposition can be used to construct control variates with nice theoretical properties, see \cite{belomestny2018stratified}.
Note that in order to compare different  variance reduction approaches, one has to analyze their complexity, that is, the number of numerical operations required to achieve a prescribed magnitude of the resulting variance.

@@

Let \((X^x_{p})_{p\geq 0}\) be a homogeneous Markov chain of the form \eqref{eq:ula_new} with $\Phi_l = \Phi,$ $l\geq 1$. Then for all $p \in \nset,$ $q \leq p$, $j < q \leq p$, $f \in \mathcal{F}_{2,\text{hom}}^p$ and $x \in \rset^d$ it holds in \(L^2(\mathsf{P})\)
f(X^x_{q})=\mathsf{E}\left[\left.f(X^x_{q})\right|\mathcal G_{j}\right]+\sum_{k=1}^{\infty}\sum_{l=j+1}^{q}\bar a_{q-l,k}(X^x_{l-1})\phi_k\left(\xi_{l}\right)
where for all $y \in \rset^d$
\bar a_{r,k}(y)=\mathsf{E}\left[f(X^y_{r})\phi_k\left(\xi_{1}\right)\right], \quad r, k\in \mathbb{N}.
\section{Variance reduction}

@@

@@

\widetilde{A}_{q,k} \in \argmin_{\psi \in \Psi} \sum_{s = 1}^{T} \left| \sum_{r = 0}^{q}  f(X_{r+l_0}^{(s)})\phi_k(\xi^{(s)}_{l_0+1}) - \psi(X^{(s)}_{l_0})\right|^2

@@

for some $l_0$ with $l_0+q\leq N+n,$

@@

\PE[A^2_{l,k}(X^x_{N+l-1})]\leq e^{\varkappa n\gamma^2}R^2_k ,\quad k=1,2,\ldots, \quad l=1,\ldots,n,

@@

for a sequence  $(R_k)$ satisfying $\sum_{k=1}^\infty R^2_k\leq C$ with constant $C$  not depending  on $\gamma$ and $n$ (see \eqref{eq:Aqk2}).  Hence if we take $W_k=R_k$ as truncation parameters in \eqref{eq:trunc}, then $\widehat A_{n-l,k}$ converges to 
\bar A_{q,k}=\argmin_{\psi\in\Psi}\mathsf{E}[|A_{q,k}(X_L)-\psi(X_L)|]^2
 as $T\to \infty,$ see Section~11 in \cite{gyorfi2006distribution}.  Moreover,

@@

\PVar[\widehat \pi_{K,n}^N(f)|\TrainSet]\leq \frac{2Ce^{\varkappa n\gamma^2}}{n}. 

@@

 Note that the above estimate is rather rough as we ignore here the fact that the difference between $\widehat A_{n-l,k}$ and $A_{n-l,k}$ is small.
If the class of functions $\Psi$ is a linear one of dimension $\mathtt{D}_\Psi,$ then the cost of computing $\widehat \pi_{K,n}^N(f)$ is of order $ \mathtt{D}_\Psi K T^2 n$. At the same time the variance of the standard estimate $ \pi_{n}^N(f)$ is of order $1/(n\gamma)$ and this bound can not be improved, see Remark~\ref{rem:var-low}. Thus, the ratio of cost-variances is of order

@@

\frac{\mathrm{cost}(\pi_{K,n}^N) \PVar[\widehat \pi_{K,n}^N(f)|\TrainSet]}{\mathrm{cost}(\pi_{n}^N)\PVar[\pi_{n}^N(f)]}\leq C e^{\varkappa n\gamma^2} K \gamma \, T^2 \mathtt{D}_\Psi. 

@@

Thus, for any fixed $K\geq 1,$  our variance reduction method is advantageous if $\gamma=\min\{1/(\mathtt{D}_\Psi T^2),1/\sqrt{n}\}$  

@@

By $H_k$, $k\in\mathbb N_0$,
we denote the normalized Hermite polynomial on $\mathbb R$, that is,
$$
H_k(x)=\frac{(-1)^k}{\sqrt{k!}}\rme^{x^2/2}\frac{\partial^k}{\partial x^k}\rme^{-x^2/2},
\quad x\in\mathbb R.
$$
For a multi-index $\mathbf{k}=(k_i)\in\mathbb N_0^d$,
$\mathbf{H}_\mathbf{k}$ denotes the normalized Hermite polynomial on $\mathbb R^d$, that is,
$$
\mathbf{H}_\mathbf{k}(\mathbf{x})=\prod_{i=1}^d H_{k_i}(x_i),\quad \mathbf{x}=(x_i)\in\mathbb R^d.
$$
In what follows, we also use the notation
$|\mathbf{k}|=\sum_{i=1}^d k_i$ for $\mathbf{k}=(k_i)\in\mathbb N_0^d$,
and we set $\mathcal G_p=\sigma(Z_1,\ldots,Z_p)$, $p\in\mathbb N$, and $\mathcal G_0=\mathrm{triv}$.
Given $N$ and $n$ as above, for $K\in\mathbb N$, denote
\begin{align}
M_{K,n}^N(f) & =\sum_{\mathbf k\colon 0<\|\mathbf{k}\|\le K}\frac{1}{n}\sum_{l=1}^{n}A_{n-l,\mathbf{k}}(X_{N+l-1})\mathbf{H}_\mathbf{k}(Z_{N+l})
\label{eq:29032018a5}
\notag
\end{align}
with \(\|\mathbf{k}\|=\max_{i} k_i\) and 
\begin{eqnarray*}
A_{q,\mathbf{k}}(y)=\sum_{r=0}^{ q } \bar{a}_{r,\mathbf{k}}(y).
\end{eqnarray*}
For an estimator $\rho(f)\in\{\pi_n^N(f),\pi_{K,n}^N(f)\}$
of $\pi(f)$ (see \eqref{eq:29032018a2} and \eqref{eq:29032018a3}), we shall be interested in the Mean Squared Error (MSE),
which can be decomposed as the sum of the squared  bias
and the  variance:
\begin{equation}
\label{eq:29032018a4}
\mathrm{MSE}\left[\rho(f)\right] = \PE\left[\left\{\rho(f)-\pi(f)\right\}^2\right]
=\left\{\PE[\rho(f)]-\pi(f)\right\}^2 +\PVar[\rho(f)].
\end{equation}
Our analysis is carried out under the following two assumptions:
\begin{description}
\item[(H1)][\textbf{Lipschitz continuity}] The potential $U$ is differentiable and $\nabla U$ is Lipschitz, that is, there exists $L < \infty$ such that
\begin{eqnarray*}
\| \nabla U(x)-\nabla U(y) \| \leq L \|x-y\|, \quad x,y\in \rset^d.
\end{eqnarray*}
\item[(H2)][\textbf{Convexity outside a ball}] There exist $K>0$ and $m>0$ such that for any $x\not\in B_{K}(0)$  it holds
\begin{equation*}
\ps{D^2U(x)}{x} \geq m\|x\|^2.
\end{equation*}
\end{description}
Let $\pi$ be the probability measure on $\rset^d$ with density $\pi(x)$ of the form \eqref{eq:stationary_distr};
for $\gamma>0$, define the Markov kernel $Q_\gamma$ associated to one step of the ULA algorithm by
\begin{equation}
\label{eq:ula_kernel}
\kerULA_{\gamma}(x,A)=\int_{A} \frac{1}{(2 \pi \gamma)^{d/2}} \text{ exp} \left\{ -\frac{1}{2\gamma} \left\| y - x + \gamma \nabla U(x)\right\|^2\right\} \rmd y
\end{equation}
for any $A \in \mathcal{B}(\rset^d)$. Note that for a homogeneous Markov chain $\left(X^x_{p}\right)_{p \in \nset_0}$ of the form \eqref{eq:chain} with constant step size $\gamma$, $\mathsf{P}(X_n^x \in A) = \kerULA^n_{\gamma}(x,A)$. 
Due to the martingale transform structure
of $M_{K,n}^N(f)$,
we have
\[
\mathsf E\left[M_{K,n}^N(f)\right]=0.
\]
Hence both estimators
$\pi_n^N(f)$ and $\pi_{K,n}^N(f)$ have the same  bias. Under assumptions {\bf (H1)} and {\bf (H2)}, the corresponding Markov chain has a unique stationary distribution $\pi_\gamma$, which is different from $\pi$. From \cite[Theorem~10]{durmus:moulines:2017} and Lemma~\ref{lem:v_ergodicity} it follows that
\begin{equation}
\label{eq:bias_2}
\tvnorm{\pi - \pi_\gamma} < C_1\sqrt{\gamma}.
\end{equation}
Let us now derive an upper bound for the variance of the classical estimator~\eqref{eq:29032018a2} for ULA-based chain.

\begin{lem} 
\label{lem:variance}
Let $f$ be a bounded Borel function and $(X^x_{p})_{p \in \nset_0}$ be a Markov chain generated by ULA with constant step size $\gamma$, satisfying {\bf (H1)} and {\bf (H2)}. Assume also that $n\gamma > C_0$ for some fixed $C_0,$ then it holds
\label{eq:var-mc}
\PVar[\pi_n^N(f)]\lesssim \frac{V(x)}{n\gamma} \eqsp,
with $V(x) = 1+ \|x\|^2$ where $\lesssim$ stands for inequality up to a constant depending on constants $L,$ $m,$  $M$ and $\mathrm{osc}(f)$.
\begin{remark}
\label{rem:var-low}
The bound in Lemma~\ref{lem:variance} is sharp and can not be improved even in the case of a normal distribution $\pi.$
Indeed,  ULA for the standard normal distribution takes form
X_{k+1} = (1-\gamma)X_k + \sqrt{2\gamma}\xi_{k+1},\quad \xi_{k+1} \sim \mathcal{N}(0,1)
Thus, setting $X_0 = 0$, we obtain
\PVar_{0} \pi_n^0(f) &=& \PVar_{0}\left[\frac{1}{n}\sum\limits_{i=1}^{n}X_i\right] 
= \frac{1}{n^2}\PVar\left[\sum\limits_{i=1}^{n}\sum\limits_{j=1}^{i}\sqrt{2\gamma}(1-\gamma)^{i-j}\xi_{j}\right]
\geq  \frac{2}{n\gamma} - \frac{4(1-\gamma)}{n^2\gamma^2}.
\end{remark}
One of the main results of this paper is the following  upper bound for the functions $A_{q,\mathbf{k}}.$

@@

Fix some $K\geq 1$ and suppose additionally that a bounded function $f$ and $\mu=\nabla U$ are $K \times d $ times continuously differentiable

@@

 Set $V(x) = 1 + \|x\|^2,$ then  it holds for $q=1,\ldots,n,$ and all $x$

@@

A^2_{q,\mathbf{k}}(x)\leq C_Ke^{\varkappa n\gamma^2}(1+V(x)),\quad 1\leq |\mathbf{k}|\leq K,
 and 
\label{eq:Aqk2sum}
\sum_{\|\mathbf{k}\|\geq K+1} A^2_{q,\mathbf{k}}(x)\leq C_{K}e^{\varkappa n\gamma^2}\sum_{I\subseteq\{1,\ldots,d\},\, I\neq \emptyset}

@@

with some positive constants $\varkappa, C_K$ not depending on $n$ and  $\gamma.$  The sum in \eqref{eq:Aqk2} runs over all nonempty subsets $I$ of the set $\{1,\ldots,d\}$ 

@@

\PVar\left(\pi_{K,n}^{N}(f)\right)\lesssim n^{-1} \gamma^{K-1}  e^{\varkappa n\gamma^2}

@@

where $\lesssim$ stands for inequality up to a constant not depending  on $n$ and $\gamma$.

@@

A_{q,\mathbf{k}}(x)=\gamma^{|\mathbf{k}'|/2}\frac{\sqrt{(\mathbf{k}-\mathbf{k}')!}}{\sqrt{\mathbf{k}!}}\mathsf{E}\left[ \partial_{Z_1}^{\mathbf{k}'} F(x,Z_1,\ldots,Z_q) \mathbf{H}_{\mathbf{k}-\mathbf{k}'}(Z_{1})\right],

@@

stands for a weak partial derivative of the functional $F_s$ that also can be viewed as discretised version of Malliavin derivative. 
Now by taking $\mathbf{k}'=\mathbf{k},$ we get

@@

A^2_{q,\mathbf{k}}(x)\leq \gamma^{|\mathbf{k}|} \PVar \left( \sum_{p=1}^{q}\partial_{Z_{1}}^{\mathbf{K}_I}f\left(X_{p}^x\right)
\right).
Also from \eqref{eq:Ask-part} we can derive
\sum_{\|\mathbf{k}\|\geq K+1} A^2_{q,\mathbf{k}}(x)\leq \sum_{I\subseteq\{1,\ldots,d\},\, I\neq \emptyset}

@@

@@

\right)\leq C_{K}\gamma^{-1}e^{\varkappa n\gamma^2}(1+V(x)),\quad q=1,\ldots,n,

@@

for some positive constants $\varkappa, C_K$ not depending on $n$ and  $\gamma.$ 

@@

In this section we illustrate the performance of the proposed variance reduction method for ULA.
First we note that from  numerical point of view another representation of the coefficients \(a_{p,l,k}\)  turns out to be more useful.
\begin{prop}
Let $q\geq l, k\in \nset$. Then the coefficients \(a_{q,l,k}(x)\) in \eqref{eq:coeff_mart}  can be alternatively represented as
a_{q,l,k}(x)=\mathsf{E}\left[\phi_k\left(\xi\right)Q_{q,l}\left(\Phi_l(x,\xi)\right)\right]
with \(Q_{q,l}(x)=\mathsf{E}\left[f(X^x_{l,q})\right],\) \(q\geq l.\)
In the case $\Phi_l=\Phi$ in \eqref{eq:ula_new} for all $l\geq 1,$ we have
\label{eq:definition-bar-a}
\bar a_{r,k}(x)=\mathsf{E}\left[\phi_k\left(\xi\right)Q_{r}\left(\Phi(x,\xi)\right)\right]
with  \(Q_{r}(x)=\mathsf{E}\left[f(X^x_{r})\right],\) $r\in \mathbb{N}.$
\end{prop}
 We construct a polynomial approximation  for each \(Q_r(x)\) in \eqref{eq:definition-bar-a} in the form:

@@

\widehat{Q}_{r}(x) = \sum_{\|\mathbf{s}\|\leq m} \widehat\beta_{\mathbf{s}} x^{\mathbf{s}},\quad s=(s_1,\ldots,s_d) \in \nset_0^d.

@@

The coefficients \(\widehat \beta_{\mathbf{s}}\in \mathbb{R}\) are obtained using a modified least-squares criteria based on $T$ independent training trajectories \(\bigl\{\bigl(X_1^{(s)},\ldots,X_{N+n}^{(s)}\bigr)\bigr\}_{s=1}^{T}\). More precisely  we define $\widehat{Q}_0(x) = f(x)$ and

@@

\widehat{Q}_r = \argmin_{\psi \in \Psi_m} \sum_{s = 1}^{T}  \left| f(X_{l_0+r}^{(s)}) - \psi(X_{l_0}^{(s)})\right|^2

@@

for \(1 \leq l_0+r \leq n-1,\) where $\Psi_m$ is a class of polynomials 
 \psi(x) = \sum_{\|\mathbf{s}\|\leq m} \alpha_{\mathbf{s}} x^{\mathbf{s}}, \quad \alpha_{\mathbf{s}}\in \mathbb{R}.

@@

\widehat a_{r,\mathbf{k}} (x) &=&\CPE{\mathbf{H}_\mathbf{k}(\xi) \widehat{Q}_{r}(x - \gamma \mu(x) + \sqrt{\gamma}\xi)}{\TrainSet}
